---
title: "Analysis of Craft Beers across the United States"
author: "David Churchman"
date: "June 26, 2017"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

#Introduction

In recent decades, the craft beer market in the United States has greatly diversified, with hundreds of different breweries all across the country brewing thousands of types of craft beers. As breweries have proliferated, states have developed regional tastes and breweries have reacted by specializing.  This report conducts an analysis of 2410 US craft beers brewed in 558 breweries across all 50 states and the District of Columbia, investigating the median bitterness and alcohol content of beers by state and the relationship between bitterness and alcohol in beers.

The "Beers" data set contains information about 2410 US craft beers brewed in 558 breweries. Below is listed the first six beers in the data set, showing information about the Alcohol by Volume (ABV), International Bitterness Unit (IBU), style, and serving size, along with a Brewery ID which will be used to link this data to the "Brewery"" data.
```{r}
#Required packages:
library(ggplot2)
library(fiftystater)
#Read in the Beers data
beers <- read.csv('beers.csv')
head(beers)
```

Additionally, below is the first six breweries in the "Breweries" data set, with information about the 558 breweries found in the Beers data set, listing the breweries by city and state, and with the common Brew_ID variable from the Beers data set.
```{r}
#Read in the breweries data
breweries <-read.csv('breweries.csv')
head(breweries)
```
#Methods and Results
The following table is the number of craft breweries in each state. The table shows low-population states like the Dakotas and West Virginia only containing one brewery, and Colorado jumps out as a high outlier with 47 breweries.
```{r}
# Generate a table displaying the number of breweries in each state.
table(breweries$State)
```

It is common to hear that craft breweries are most popular in the Pacific Northwest and the Midwest, but it is difficult to verify this in the above table, so below is a map of the number of breweries in each state, where the states with the most breweries are bright red, and the states with the fewest breweries are a deep blue.  From the map, it is clear that there is a cluster of states on the west coast with the most breweries, but the bright red Colorado and Texas are notable exceptions in the middle of the mostly blue center of the country. There is also a cluster in the Midwest and Northeast of states with more breweries. Unsurprisingly, the least craft breweries are found in the least populated parts of the United States in the middle of the country and the south.
```{r}
source('CreateMap.R', print.eval = TRUE)
```
Having established where craft breweries are most common, this report will proceed to analyze differences in state tastes for bitterness and alcohol content in craft beers.  In order to get more information about the individual beers brewed in each state, the below code merges the "Beers" and "Breweries" data sets, cross-referencing them using the common element of Brewery ID.
```{r}
#Merge the 2 raw data files and do some basic cleaning
source('BeerMerge.R')
```

The new data set has information about individual beers brewed in each state. As an example of the data available, below are the first six beers in the combined data set.
```{r}
head(beermerge)
```
And the last six beers:
```{r}
tail(beermerge)
```
In the above rows, it is apparent that the data set does not have complete information for every element. "NA" represents information that was not recorded in the data set. Below is a table of the number of "NAs" in each variable.
```{r}
#Number of NAs in each variable
colSums(is.na(beermerge))
```
Almost half, 1005, of the beers have missing information about their IBU, and a few dozen, 62, have missing information regarding ABV. It is likely that some craft breweries do not know or are not required to know the IBU of their beers, and there may be incomplete data gathering around ABV.

The International Bitterness Unit (IBU) is a measure of bitterness for beers. The way hops are prepared and brewed into the beer can have a large impact on the flavor, with some beers like India Pale Ales tasting very bitter, and hence scoring a higher IBU, and some beers like lagers not tasting bitter at all, scoring a low IBU. Breweries in different states cater to different tastes of bitterness. Below is the median IBU of craft beers by state.
```{r}
#Median IBU by state
stateIBU<-aggregate(IBU~State, beermerge,median, na.action=na.omit)
stateIBU
```

Below is a bar graph of the above table, which makes it easier to see that Maine has the highest median beer IBU, and West Virginia is close behind. West Virginia, however, has only has one brewery, so this is probably not representative of the tastes of most West Virginians. A surprising low outlier is Wisconsin, as Wisconsin is a state know for its beers, but these beers must be milder or less bitter than in other states known for their beers like Colorado. This graph also shows that most states have a median IBU between 20 and 50, which suggests a maximum threshold for bitterness. 
```{r}
#Barplot of IBU by state
ggplot(data=stateIBU, aes(x=State, y=IBU, fill=IBU)) +
    scale_fill_distiller(palette="Reds") +
    geom_bar(stat="identity") +
    theme(text = element_text(size=10),
          axis.text.x = element_text(angle=90, hjust=1,vjust=.5))
```

Similarly, the level of alcohol content, measured by Alcohol by Volume (ABV), varies from state to state. Below is a table of the median alcohol content of craft beers by state.
```{r}
#Median ABV by state
stateABV<-aggregate(ABV~State, beermerge, median, na.action = na.omit)
stateABV
```

The below bar graph also shows the median ABV by state. Utah is a low outlier, which may have roots in the influence of the Mormon church in Utah and their prohibitions against alcohol. Kentucky and the District of Columbia have the highest median ABV, followed closely by Maine, New Mexico and West Virginia.
```{r}
#BarPlot of ABV by state
ggplot(data=stateABV, aes(x=State, y=ABV, fill=ABV)) +
    geom_bar(stat="identity") +
    theme(text = element_text(size=10),
          axis.text.x = element_text(angle=90, hjust=1,vjust=.5))
```

Maine appearing as a high outlier in both median IBU and ABV suggests that there may be some relationship between IBU and ABV. The below plot shows that there is indeed a positive linear relationship between the two, where the higher the bitterness of a beer, the higher its alcohol content. However, there also appears to be a maximum threshold ABV of about 0.10 that most beers do not cross, with two high outliers. This might be a limitation of the brewing process, or it might be that an ABV greater than 0.10 is too alcoholic and and has an unpalatable flavor. IBU also seems to be clustered, with a significant drop-off after an IBU of 50 and even less cpmmon after 100.
```{r}
#ABV vs. IBU scatterplot
ggplot(data=beermerge,aes(IBU,ABV)) +
    geom_point(color="blue") +
    geom_smooth(method=lm,color="red")
```

Out of curiosity, those high ABV outliers are a Quadrupel from Colorado, and an English Barleywine from Kentucky, both sporting an alcohol content more usually associated with wine than beer.
```{r}
#Finding ABV >0.12 for high outliers
beermerge[which(beermerge$ABV > 0.12),c(2,4,6, 8,9,10)]
```


#Conclusion

Brewery data from all fifty states and the District of Columbia show a large variety of craft beers from many breweries. Breweries are most frequent on the coasts, though all states have at least one craft brewery. There are a variety of regional tastes in alcohol content and bitterness around the country, but in general they fall within similar ranges. There is a correlation between more bitter beers and higher alcohol content, though, again, most beers fall within a fairly tight cluster of values for both. Altogether, the large number of craft breweries and the many craft beers they create is a diverse market, catering to a large variety of tastes across the United States.